---
title: "US Storm - Damage and Casualty Analysis"
author: "Shashi Ponraja"
date: "17/03/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

### Abstract

The following is an analysis of data collected from NOAA, examining the impact of the various storm types in USA.

## Data Pre-Processing

The data was downloaded directly from the assignment URL, using the following code

```{r}
fileURL = 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
if (!file.exists('./data/NOAA_Storm_Data_decompressed.csv')){
  download.file(fileURL, destfile = "./data/NOAA_Storm_Data.csv.bz2", method="curl")
  storm_data = read.csv('./data/NOAA_Storm_Data.csv.bz2') 
  write.csv(storm_data,'./data/NOAA_Storm_Data_decompressed.csv')
}else{
  storm_data = read.csv('./data/NOAA_Storm_Data_decompressed.csv')
}

```

Subsequent data processing was performed 'on-the-fly' from the original source data - see related code below the final analyses.

### Results

## Deaths

The 5 types of storm events that are responsible for the highest total deaths are -

```{r}
fatalities_by_eventType = aggregate(FATALITIES ~ EVTYPE, data = storm_data, FUN=sum)

ordered_fatalities_by_eventType = fatalities_by_eventType[order(-fatalities_by_eventType$FATALITIES),]
ordered_fatalities_by_eventType[1:5,]
```

The 5 types of storm events that, on average, cause the most fatalities are -

```{r}
average_fatalities_by_eventType = aggregate(FATALITIES ~ EVTYPE, data = storm_data, FUN=mean)

ordered_avg_fatalities_by_eventType = average_fatalities_by_eventType[order(-average_fatalities_by_eventType$FATALITIES),]
ordered_avg_fatalities_by_eventType[1:5,]
```

## Casualties

The 5 types of storm events that, in total, have lead to the most casualties are -

```{r}
injuries_by_eventType = aggregate(INJURIES ~ EVTYPE, data = storm_data, FUN = sum)
ordered_injuries_by_eventType = injuries_by_eventType[order(-injuries_by_eventType$INJURIES),]
ordered_injuries_by_eventType[1:5,]

```

The 5 types of storm events that, on average, cause to the most casualties are -

```{r}
avg_injuries_by_eventType = aggregate(INJURIES ~ EVTYPE, data = storm_data, FUN = mean)
avg_ordered_injuries_by_event_Type = avg_injuries_by_eventType[order(-avg_injuries_by_eventType$INJURIES),]
#avg_ordered_injuries_by_event_Type[1:5,]
#par(las=2)
barplot(avg_ordered_injuries_by_event_Type$INJURIES[1:5], main="Average Casualties for top 5 storm types", ylab = "Casualties",xlab="Storm Type",space = rep(0.0,5),cex.names = 0.5,names.arg = levels(factor(avg_ordered_injuries_by_event_Type$EVTYPE[1:5])))

```

## Cost of Damage to Property and Crops

```{r, echo=FALSE}
## clean the data
exponenter = function(x){
  if ((x == 'h') || (x == 'H') ){
    return(2)
  }
  if ((x == 'm') || (x == 'M') ){
    return(6)
  }
  
  if ((x == 'K')){
    return(3)
  }
  if ((x == '') || (x == '-') || (x == '?') || (x == '+')){
    return(0)
  }
  return(x)
}
storm_data$PROPDMGEXP_cleaned = sapply(storm_data$PROPDMGEXP, exponenter)
storm_data$CROPDMGEXP_cleaned = sapply(storm_data$CROPDMGEXP, exponenter)

storm_data$DollarValue_CROPDMG = storm_data$CROPDMG * 10 ^ storm_data$CROPDMGEXP_cleaned
storm_data$DollarValue_PROPDMG = storm_data$PROPDMG * 10 ^ storm_data$PROPDMGEXP_cleaned

Total_CROPDMG_by_eventType = aggregate(DollarValue_CROPDMG ~ EVTYPE, storm_data, FUN = sum)
Total_PROPDMG_by_eventType = aggregate(DollarValue_PROPDMG ~ EVTYPE, storm_data, FUN = sum)

Avg_CROPDMG_by_eventType = aggregate(DollarValue_CROPDMG ~ EVTYPE, storm_data, FUN = mean)
Avg_PROPDMG_by_eventType = aggregate(DollarValue_PROPDMG ~ EVTYPE, storm_data, FUN = mean)

```


The 5 storm types that cumulatively cause the most crop damage are -

```{r}
ordered_Total_CROPDMG_by_eventType = Total_CROPDMG_by_eventType[order(-Total_CROPDMG_by_eventType$DollarValue_CROPDMG),]
ordered_Total_CROPDMG_by_eventType[1:5,]

```

and on average - 

```{r}
ordered_Avg_CROPDMG_by_eventType = Avg_CROPDMG_by_eventType[order(-Avg_CROPDMG_by_eventType$DollarValue_CROPDMG),]
ordered_Avg_CROPDMG_by_eventType[1:5,]
```

The 5 storm types that, in total, are responsible for the most property damage -

```{r}
ordered_Total_PROPDMG_by_eventType = Total_PROPDMG_by_eventType[order(-Total_PROPDMG_by_eventType$DollarValue_PROPDMG),]
ordered_Total_PROPDMG_by_eventType[1:5,]
```

and on average -

```{r}
ordered_Avg_PROPDMG_by_eventType = Avg_PROPDMG_by_eventType[order(-Avg_PROPDMG_by_eventType$DollarValue_PROPDMG),]
ordered_Avg_PROPDMG_by_eventType[1:5,]
```
